E-graphs With Bindings

### 2025-01-15 LICS_2025/e_graphs_with_bindings/E-graphs-with-bindings.tex
wip
intro fix

\documentclass[peerreviewcls]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
%Template version as of 6/27/2024

@@

E-graphs were first introduced in the automated theorem proving literature in the 1980s~\cite{nelson1980techniques}, but recent years have seen renewed interest particularly for optimisation and synthesis using so called \emph{equality saturation}~\cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}, the repeated application of rewrites to an e-graph until a fixpoint is reached. 
The wide range of application domains led to extensions of the e-graph formalism, for instance to account for conditional rewriting~\cite{singher2023colored} or  to combine e-graphs with database queries (\emph{relational e-matching})~\cite{zhang_relational_2022}.
Of particular interest to us is using e-graphs in the context of functional programming languages, which requires a representation of the $\lambda$-calculus using it. 
Pioneering work in this direction has been done by~\cite{koehler2022sketchguided}.
The well-known twin challenges of binding and capture-avoiding substitution remain however somewhat problematic in this work because DAN: IT IS UNCLEAR WHY THIS IS PROBLEMATIC IN LOC. CIT. THE TEXT BELOW IS UNCONVINCING. CAN YOU ILLUSTRATE THE PROBLEM WITH SOME SIMPLE EXAMPLES? 
\begin{quotation}
There are domains, however, where the effective usage of e-graphs requires \textit{ad hoc} tweaks to achieve the desired performance, namely domains that include variable \textit{binding} and the closely related notion of \textit{substitution}, for example, lambda calculus with $\beta$-reduction.
The first issue appears due to the fact that lambda calculus terms are quotiented by renaming of bound variables.
If such renaming is not accounted for by the language encoding, \textit{e.g.,} by using \textit{de Brujin} indices, populating an e-graph with $\alpha$-equivalent terms increases the size of an e-graph without adding any meaningful equivalences.
The other issue is that having binding entails having a substitution and computing substitution in e-graphs is hard as it is not a mere substitution of terms but rather a substitution of equivalence classes that represent a lot of terms at once.
\end{quotation}

@@

\section{Categorical semantics}

Free semilattice enrichment of a closed SMC.

>>>>

\documentclass[peerreviewcls]{IEEEtran}
% \IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
%Template version as of 6/27/2024

@@

E-graphs were first introduced in the automated theorem proving literature in the 1980s~\cite{nelson1980techniques}, but recent years have seen renewed interest particularly for optimisation and synthesis using so called \emph{equality saturation}~\cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}, the repeated application of rewrites to an e-graph until a fixpoint is reached. 
The wide range of application domains led to extensions of the e-graph formalism, for instance to account for conditional rewriting~\cite{singher2023colored} or  to combine e-graphs with database queries (\emph{relational e-matching})~\cite{zhang_relational_2022}.
Of particular interest to us is using e-graphs in the context of functional programming languages, which requires a representation of the $\lambda$-calculus using it. 
\textcolor{red}{Pioneering work in this direction has been done by~\cite{koehler2022sketchguided} where
the authors outlined the two interconnected challenges of binding and capture-avoiding substitution that arise when doing equality saturation for $\lambda$-calculus with e-graphs.
As e-graph endeavours to represent multiple terms at once with as much sharing as possible, using variable names for binding means that two $\alpha$-equivalent terms can not be shared as they are structurally different.
Even though such simple encoding makes expressing $\beta$-reduction as a rewrite rule in terms of explicit substitution simpler, applying such rules requires computing extra free variable predicates and populates a given e-graph with extra nodes for explicit substitution.
On the other hand, encoding binding using $\textit{de Brujin}$ while still fails to represent $\alpha$-equivalent terms as being structurally equivalent, as \textit{e.g.}, in $\lambda . \%0 ((\lambda . \%1 \%0) 42)$ the same binder is represented with $\%0$ and $\%1$, avoids explicit substitution when implementing $\beta$-reduction at the cost of including \textit{ad hoc} operators for index shifting into the rewrite rules.
Generally, the incurred redundancy these challenges remain problematic because variable binding is not inherent to e-graphs as they do not provide an internal support for it.
Other works suggested an approach to add internal support for binding which however made the formulation of e-graph considerably more complicated~\cite{slotted-egraphs}.
}
\textcolor{blue}{The reference above is a workshop unpublished work, I do not know if it is good to use such reference.} 
% The well-known twin challenges of binding and capture-avoiding substitution remain however somewhat problematic in this work because DAN: IT IS UNCLEAR WHY THIS IS PROBLEMATIC IN LOC. CIT. THE TEXT BELOW IS UNCONVINCING. CAN YOU ILLUSTRATE THE PROBLEM WITH SOME SIMPLE EXAMPLES? 
% \begin{quotation}
% There are domains, however, where the effective usage of e-graphs requires \textit{ad hoc} tweaks to achieve the desired performance, namely domains that include variable \textit{binding} and the closely related notion of \textit{substitution}, for example, lambda calculus with $\beta$-reduction.
% The first issue appears due to the fact that lambda calculus terms are quotiented by renaming of bound variables.
% If such renaming is not accounted for by the language encoding, \textit{e.g.,} by using \textit{de Brujin} indices, populating an e-graph with $\alpha$-equivalent terms increases the size of an e-graph without adding any meaningful equivalences.
% The other issue is that having binding entails having a substitution and computing substitution in e-graphs is hard as it is not a mere substitution of terms but rather a substitution of equivalence classes that represent a lot of terms at once.
% \end{quotation}

@@

\input{sections/categorical_semantics}

### 2025-01-15 LICS_2025/e_graphs_with_bindings/sections/categorical_semantics.tex
intro fix

A set of $\Sigma$-terms generated by monoidal signature $\Sigma$ is given by the following grammar
\[
\tau \Coloneq \phi \;|\; \id_{I} \;|\; \id_{A} \;|\; \sym_{A,B} \;|\; \tau_{1};\tau_{2} \;|\; \tau_{1} \otimes \tau_{2}
\],
where $\phi$ ranges over the set of generators $\Sigma_{M}$ and $A,B$ range over the set of objects $\Sigma_{O}$.
\end{definition}

@@

In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice enriched $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
Being a 2-functor it also lifts an adjunction in $\catname{Cat}$ $F \dashv G$ into an $\catname{Slat}$-adjunction.
Thus, applying the functor to $\catname{Sc}(\Sigma)$ we get a closed symmetric monoidal $\catname{Slat}$-category $\catname{Sc}(\Sigma)^{+}$ by lifting the adjunction $- \otimes B \dashv B \xRightarrow{} -$.

>>>>

A set of $\Sigma$-terms generated by monoidal signature $\Sigma$ is given by the following grammar
\[
\tau \coloneq \phi \;|\; \id_{I} \;|\; \id_{A} \;|\; \sym_{A,B} \;|\; \tau_{1};\tau_{2} \;|\; \tau_{1} \otimes \tau_{2}
\],
where $\phi$ ranges over the set of generators $\Sigma_{M}$ and $A,B$ range over the set of objects $\Sigma_{O}$.
\end{definition}

@@

In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice enriched $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
Being a 2-functor it also lifts an adjunction in $\catname{Cat}$ $F \dashv G$ into an $\catname{Slat}$-adjunction.
Thus, applying the functor to $\catname{Sc}(\Sigma)$ we get a closed symmetric monoidal $\catname{Slat}$-category $\catname{Sc}(\Sigma)^{+}$ by lifting the adjunction $(- \otimes B, B \xRightarrow{} -, \eta, \varepsilon)$.
We can similarly derive $\catname{SLat}$-enriched versions of  $\Lambda_{A,B,C}$ and $\textbf{ev}_{A,B}$ morphisms using the lifted adjunction as $\textbf{ev}_{A,B} = \varepsilon_{A}$ and $\Lambda_{A,B,C}(f) = \eta; (B \xRightarrow{} f)$.
The following equation can then be derived which will be useful in encoding equivalences inside abstractions.
\begin{align*}
\Lambda_{A,B,C}(f + g) &= \eta;(B \xRightarrow{} (f + g))\\
                       &= \eta;(B \xRightarrow{} (f) + B \xRightarrow{} (g))\\
                       &= \eta;(B \xRightarrow{} (f)) + \eta;(B \xRightarrow{} (g))\\
                       &= \Lambda_{A,B,C}(f) + \Lambda_{A,B,C}(g)
\end{align*}
